Subsetting plant species list by fruiting species

### 2024-05-29 code/L1/L1_P3_TropicalAndes_plant_fruiting_subset.Rmd
inital commit

---
title: "Plant trait cleaning and combining"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script subsets plant trait data by synonyms species list and combine."
data input: "ADD"
data output: "ADD"
date: "2023-11-04"
output: html_document
---

@@

```{r}
TRY_traits %>%
  count(TraitName)
```

@@

Final trait list:
- dispersal syndrome
- Fruit mass
- Fruit length
- Fruit type
- Plant growth form
- Plant height
- Plant Lifespan
- Seed mass
- Seed length
- Seed width

# Standardize trait measurements
## Create functions
```{r}
# Numeric data
numeric_data_average <- function(data, unit, traitname) {
# remove rows if TraitValue is non-numeric
data <- data %>%
  filter(!is.na(as.numeric(TraitValue))) %>%
  mutate(TraitValue = as.numeric(TraitValue))
# remove duplicate rows
data <- data %>%
  distinct(.keep_all = TRUE)
# calculate average value per trait per species
average_data <- aggregate(TraitValue ~ Accepted_species, data, FUN = mean, na.rm = TRUE)
# Remove rows with missing or empty species names
average_data <- average_data[!is.na(average_data$Accepted_species) & average_data$Accepted_species !="", ]
# Add unit & trait column
average_data$Unit <- unit
average_data$TraitName <- traitname

return(average_data)
}
```

```{r}
# categorical data
factor_data_merge <- function(data, traitname) {
# remove duplicate rows
data <- data %>%
  distinct(.keep_all = TRUE)
# remove rows with missing or empty species names
data <- data[!is.na(data$Accepted_species) & data$Accepted_species != "", ]
# Subset data to just species and TraitValue
data <- data[,c("Accepted_species", "TraitValue")]
# Find the most common trait value for each species
  most_common_values <- aggregate(. ~ Accepted_species, data, function(x) {
    # Find the mode (most common value) of the trait values
    mode_value <- names(sort(table(x), decreasing = TRUE))[1]
    return(mode_value)
  })
# Merge most common trait values back to the original data
merged_data <- merge(data, most_common_values, by = "Accepted_species", all.x = TRUE)

@@

## Dispersal syndrome
```{r}
traits_dispersal <- traits %>%
  filter(grepl('Dispersal syndrome|whole plant dispersal syndrome|dispersal_syndrome_1|dispersal_syndrome_2', TraitName))
traits_dispersal$OrigTraitValue <- traits_dispersal$TraitValue
nrow(traits_dispersal)
length(unique(traits_dispersal$Accepted_species))
```

```{r}
# remove rows containing weight or mass
traits_dispersal <- traits_dispersal[!grepl("weight|mass|air-dry|oven|mean|SE",traits_dispersal$TraitValue, ignore.case = TRUE),]
nrow(traits_dispersal)
length(unique(traits_dispersal$Accepted_species))
```

```{r}
# remove rows where TraitValue is numeric
traits_dispersal <- traits_dispersal %>%
  filter(is.na(as.numeric(TraitValue)))
nrow(traits_dispersal)
length(unique(traits_dispersal$Accepted_species))
```

```{r}
traits_dispersal %>%
  distinct(TraitValue)
```

Reclassify to the following:
anemochorous - wind dispersal
anthropochorous - human dispersal
autochorous - non-aided dispersal
hydrochorous - water dispersal
myrmecochorous - ant dispersal
zoochorous - animal dispersal
unspecialized - multiple
```{r}
traits_dispersal <- traits_dispersal %>%
  filter() %>%
  mutate(TraitValue= case_when(
    grepl('Eaten|Animal|Bird|Mammal|animals|endozoochor|dysochor|epizoochor|zoochor|Vertebrates|Reptiles+birds|Endo-zoochory|mammals|exozoochory|birds|Zoochory|vertebrate|Dys-zoochory|Ornithochory|Epi-zoochory|endozoochorous|endozoochory|ectozoochorous|zoochorous', OrigTraitValue, ignore.case = TRUE) ~ "zoochorous",
    grepl('Ants', OrigTraitValue, ignore.case = TRUE) ~ "myrmecochorous",
    grepl('wind|Anemochory|anemochor|meteorochor|chamaechor|Anemochory|anemochorous', OrigTraitValue, ignore.case = TRUE) ~ "anemochorous",
    grepl('human|agochor|man|hemerochor|anthropochorous', OrigTraitValue, ignore.case = TRUE) ~ "anthropochorous",
    grepl('No|Unassisted|autochor|blastochor|ombrochor|Autochory|Barochory|ballochor|herpochor|autochorous', OrigTraitValue, ignore.case = TRUE) ~ "autochorous",
    grepl('water|nautochor|floating|hydrochor|hydrochorous', OrigTraitValue, ignore.case = TRUE) ~ "hydrochorous",
    grepl('boleochor|combination|non specialized|unspecialized', OrigTraitValue, ignore.case = TRUE) ~ "unspecialized"
  )) %>%
  filter(!is.na(TraitValue))
nrow(traits_dispersal)
length(unique(traits_dispersal$Accepted_species))
```

```{r}
traits_dispersal %>%
  distinct(TraitValue)
```

```{r}
# standardize trait names
traits_dispersal <- traits_dispersal %>%
  mutate(TraitName = "DispersalSyndrome")
```

```{r}
averagetraits_dispersal <- factor_data_merge(traits_dispersal, "DispersalSyndrome")
```

```{r}
nrow(averagetraits_dispersal)
length(unique(averagetraits_dispersal$Accepted_species))
```

```{r}
averagetraits_dispersal %>%
  count(TraitValue)
```

```{r}
summary(averagetraits_dispersal)
```

## Fruit dry mass
```{r}
traits_fruitMass <- traits %>%
  filter(grepl('Fruit dry mass', TraitName))
traits_fruitMass$OrigTraitValue <- traits_fruitMass$TraitValue
nrow(traits_fruitMass)
length(unique(traits_fruitMass$Accepted_species))
```

```{r}
# remove non-numeric rows
traits_fruitMass <- traits_fruitMass %>%
  filter(!is.na(as.numeric(TraitValue))) %>%
  mutate(TraitValue = as.numeric(TraitValue))

nrow(traits_fruitMass)
length(unique(traits_fruitMass$Accepted_species))
```

```{r}
# check that units match
print(unique(traits_fruitMass$Unit))
```

```{r}
# standardize trait names with units
traits_fruitMass <- traits_fruitMass %>%
  mutate(TraitName = "FruitMass_mg")
```

```{r}
averagetraits_fruitMass <- numeric_data_average(traits_fruitMass, "mg", "FruitMass_mg")
```

```{r}
nrow(averagetraits_fruitMass)
length(unique(averagetraits_fruitMass$Accepted_species))
```

```{r}
summary(averagetraits_fruitMass)
```

## Fruit length
```{r}
traits_fruitLength <- traits %>%
  filter(grepl('Fruit length|maximum fruit length|minimum fruit length|fruit_length_min|fruit_length_max|fruit_length_mean', TraitName))
traits_fruitLength$OrigTraitValue <- traits_fruitLength$TraitValue

nrow(traits_fruitLength)
length(unique(traits_fruitLength$Accepted_species))
```

```{r}
# remove rows if TraitValue is non-numeric
traits_fruitLength <- traits_fruitLength %>%
  filter(!is.na(as.numeric(TraitValue))) %>%
  mutate(TraitValue = as.numeric(TraitValue))

nrow(traits_fruitLength)
length(unique(traits_fruitLength$Accepted_species))
```

```{r}
# check that units match
print(unique(traits_fruitLength$Unit))
```

```{r}
# if units = cm, multiply trait value by 10 and change units to mm
traits_fruitLength <- traits_fruitLength %>%
  mutate(TraitValue = if_else(Unit == "cm", TraitValue * 10, TraitValue),
         Unit = if_else(Unit == "cm", "mm", Unit))
```

```{r}
# standardize trait names with units
traits_fruitLength <- traits_fruitLength %>%
  mutate(TraitName = "FruitLength_mm")
```

```{r}
averagetraits_fruitLength <- numeric_data_average(traits_fruitLength, "mm", "FruitLength_mm")
```

```{r}
nrow(averagetraits_fruitLength)
length(unique(averagetraits_fruitLength$Accepted_species))
```

```{r}
summary(averagetraits_fruitLength)
```

## Plant height
```{r}
traits_plantHeight <- traits %>%
  filter(grepl('Plant height generative|Plant height vegetative|maximum whole plant height|minimum whole plant height|plant_height|plant_height_max|plant_height_min|whole plant height', TraitName))
traits_plantHeight$OrigTraitValue <- traits_plantHeight$TraitValue

nrow(traits_plantHeight)
length(unique(traits_plantHeight$Accepted_species))
```

```{r}
# remove non-numeric rows
traits_plantHeight <- traits_plantHeight %>%
  filter(!is.na(as.numeric(TraitValue))) %>%
  mutate(TraitValue = as.numeric(TraitValue))

nrow(traits_plantHeight)
length(unique(traits_plantHeight$Accepted_species))
```

```{r}
print(unique(traits_plantHeight$Unit))
```

```{r}
# standardize trait names with units
traits_plantHeight <- traits_plantHeight %>%
  mutate(TraitName = "PlantHeight_m")
```

```{r}
averagetraits_plantHeight <- numeric_data_average(traits_plantHeight, "m", "PlantHeight_m")
```

```{r}
nrow(averagetraits_plantHeight)
length(unique(averagetraits_plantHeight$Accepted_species))
```

```{r}
summary(averagetraits_plantHeight)
```

## Plant lifespan/longevity
```{r}
traits_plantLifespan <- traits %>%
  filter(grepl('Plant lifespan \\(longevity\\)|longest whole plant longevity|maximum whole plant longevity|plant_lifespan', TraitName))
traits_plantLifespan$OrigTraitValue <- traits_plantLifespan$TraitValue

nrow(traits_plantLifespan)
length(unique(traits_plantLifespan$Accepted_species))
```

```{r}
# remove rows if TraitValue is non-numeric
traits_plantLifespan <- traits_plantLifespan %>%
  filter(!is.na(as.numeric(TraitValue))) %>%
  mutate(TraitValue = as.numeric(TraitValue))

nrow(traits_plantLifespan)
length(unique(traits_plantLifespan$Accepted_species))
```

```{r}
# check that units match
print(unique(traits_plantLifespan$Unit))
traits_plantLifespan$Unit <- "years"
print(unique(traits_plantLifespan$Unit))
```

```{r}
# standardize trait names with units
traits_plantLifespan <- traits_plantLifespan %>%
  mutate(TraitName = "PlantLifespan_years")

```

```{r}
averagetraits_plantLifespan <- numeric_data_average(traits_plantLifespan, "years", "PlantLifespan_years")
```

```{r}
nrow(averagetraits_plantLifespan)
length(unique(averagetraits_plantLifespan$Accepted_species))
```

```{r}
summary(averagetraits_plantLifespan)
```

## Fruit type
```{r}
traits_fruitType <- traits %>%
  filter(grepl('Fruit type|fruit type|fruit_type', TraitName))
traits_fruitType$OrigTraitValue <- traits_fruitType$TraitValue

@@

```{r}
traits_fruitType %>%
  distinct(TraitValue)
```

```{r}
# standardize trait values
traits_fruitType <- traits_fruitType %>%
  filter() %>%
  mutate(TraitValue= case_when(
    grepl('drupe|ectozoochorous|zoochorous', OrigTraitValue, ignore.case = TRUE) ~ "zoochorous",
    grepl('capsule|Capsule|Capsule |', OrigTraitValue, ignore.case = TRUE) ~ "myrmecochorous",
  )) %>%
  filter(!is.na(TraitValue))
```

@@

## Seed mass
```{r}
traits_seedMass <- traits %>%
  filter(grepl('Seed dry mass|seed mass|seed_mass_max|seed_mass_mean|seed_mass_min', TraitName))
traits_seedMass$OrigTraitValue <- traits_seedMass$TraitValue

nrow(traits_seedMass)
length(unique(traits_seedMass$Accepted_species))
```

```{r}
# remove non-numeric rows
traits_seedMass <- traits_seedMass %>%
  filter(!is.na(as.numeric(TraitValue))) %>%
  mutate(TraitValue = as.numeric(TraitValue))

nrow(traits_seedMass)
length(unique(traits_seedMass$Accepted_species))
```

```{r}
print(unique(traits_seedMass$Unit))
```

```{r}
# convert rows with unit = g to mg; multiply by 1000
traits_seedMass <- traits_seedMass %>%
  mutate(TraitValue = if_else(Unit == "g", TraitValue * 1000, TraitValue),
         Unit = if_else(Unit == "g", "mg", Unit))
```

```{r}
# standardize trait names with units
traits_seedMass <- traits_seedMass %>%
  mutate(TraitName = "SeedMass_g")
```

```{r}
averagetraits_seedMass <- numeric_data_average(traits_seedMass, "mg", "SeedMass_g")
```

```{r}
nrow(averagetraits_seedMass)
length(unique(averagetraits_seedMass$Accepted_species))
```

```{r}
summary(averagetraits_seedMass)
```

## Seed length
```{r}
traits_seedLength <- traits %>%
  filter(grepl('Seed length|seed_length_max|seed_length_mean|seed_length_min', TraitName))
traits_seedLength$OrigTraitValue <- traits_seedLength$TraitValue

nrow(traits_seedLength)
length(unique(traits_seedLength$Accepted_species))
```

```{r}
# remove non-numeric rows
traits_seedLength <- traits_seedLength %>%
  filter(!is.na(as.numeric(TraitValue))) %>%
  mutate(TraitValue = as.numeric(TraitValue))

nrow(traits_seedLength)
length(unique(traits_seedLength$Accepted_species))
```

```{r}
print(unique(traits_seedLength$Unit))
```

```{r}
# standardize trait names with units
traits_seedLength <- traits_seedLength %>%
  mutate(TraitName = "SeedLength_mm")
```

```{r}
averagetraits_seedLength <- numeric_data_average(traits_seedLength, "mm", "SeedLength_mm")
```

```{r}
nrow(averagetraits_seedLength)
length(unique(averagetraits_seedLength$Accepted_species))
```

```{r}
summary(averagetraits_fruitLength)
```

## Seed width
```{r}
traits_seedWidth <- traits %>%
  filter(grepl('Seed width|seed_width_max|seed_width_mean|seed_width_min'
, TraitName))
traits_seedWidth$OrigTraitValue <- traits_seedWidth$TraitValue

nrow(traits_seedWidth)
length(unique(traits_seedWidth$Accepted_species))
```

```{r}
# remove non-numeric rows
traits_seedWidth <- traits_seedWidth %>%
  filter(!is.na(as.numeric(TraitValue))) %>%
  mutate(TraitValue = as.numeric(TraitValue))

nrow(traits_seedWidth)
length(unique(traits_seedWidth$Accepted_species))
```

```{r}
print(unique(traits_seedWidth$Unit))
```

```{r}
# standardize trait names with units
traits_seedWidth <- traits_seedWidth %>%
  mutate(TraitName = "SeedWidth_mm")
```

```{r}
averagetraits_seedWidth <- numeric_data_average(traits_seedWidth, "mm", "SeedWidth_mm")
```

```{r}
nrow(averagetraits_seedWidth)
length(unique(averagetraits_seedWidth$Accepted_species))
```

```{r}
summary(averagetraits_seedWidth)
```

## Growth form
```{r}
traits_growthForm <- traits %>%
  filter(grepl('Plant growth form|growth_form|whole plant growth form'
, TraitName))
traits_growthForm$OrigTraitValue <- traits_growthForm$TraitValue

nrow(traits_growthForm)
length(unique(traits_growthForm$Accepted_species))
```

```{r}
# remove numeric rows
traits_growthForm <- traits_growthForm %>%
  filter(is.na(as.numeric(TraitValue)))

nrow(traits_growthForm)
length(unique(traits_growthForm$Accepted_species))
```

```{r}
# look at unique trait values
traits_growthForm %>%
  distinct(TraitValue)
```

```{r}
averagetraits_growthForm <- factor_data_merge(traits_growthForm, "Growth Form")
```

# Combine all traits
```{r}
# add unit column to categorical traits
cat_traits <- rbind(averagetraits_dispersal, averagetraits_fruitType, averagetraits_growthForm)
cat_traits$Unit <- ""
clean_traits <- rbind(averagetraits_fruitLength, averagetraits_fruitMass, averagetraits_plantHeight, averagetraits_plantLifespan, averagetraits_seedLength, averagetraits_seedMass, averagetraits_seedWidth, cat_traits)
nrow(clean_traits)
```

```{r}
clean_traits %>%
  count(TraitName)
```

# Summary
```{r}
print("Number of records: ")
nrow(clean_traits)
print("Number of species: ")
length(unique(clean_traits$Accepted_species))
```

# Get one trait value per species
```{r}
#remove unit column from dataframe
clean_traits <- clean_traits[, !colnames(clean_traits) %in% "Unit", drop = FALSE]
```

```{r}
# separate numeric and non-numeric traits
numeric_traits <- clean_traits[clean_traits$TraitName %in% c("Fruit Length", "Fruit Mass", "Plant Height", "Plant Lifespan", "Seed Length", "Seed Width", "Seed Mass"), ]
cat_traits <- clean_traits[clean_traits$TraitName %in% c("Dispersal Syndrome", "Fruit Type", "Growth Form"), ]

numeric_traits$TraitValue <- as.numeric(numeric_traits$TraitValue)
```

```{r}
# numeric traits
## Group by species and TraitName, and calculate the mean TraitValue for each combination
numeric_traits_combined <- numeric_traits %>%
  group_by(species, TraitName) %>%
  summarise(TraitValue = mean(TraitValue, na.rm = TRUE)) %>%
  ungroup()

# non-numeric traits
cat_traits_combined <- cat_traits %>%
  group_by(species) %>%
  summarise_all(.funs = list(mode_trait = function(x) {
    mode_value <- names(sort(table(x), decreasing = TRUE))[1]  # Calculate mode
    return(mode_value)
  }))
```

```{r}
#rename columns to original values
colnames(cat_traits_combined)[which(colnames(cat_traits_combined) == "TraitValue_mode_trait")] <- "TraitValue"
colnames(cat_traits_combined)[which(colnames(cat_traits_combined) == "TraitName_mode_trait")] <- "TraitName"

# Remove rows with "?" or numbers in the TraitValue column
cleaned_cat_traits <- cat_traits_combined %>%
  filter(!grepl("\\?", TraitValue) & !grepl("\\d", TraitValue))
```

```{r}
#combine numeric and non-numeric traits
combined_clean_traits <- rbind(cat_traits_combined, numeric_traits_combined)
```

```{r}
combined_clean_traits %>%
  count(TraitName)
```

# Remove traits with less than 10 records
```{r}
combined_clean_traits <- combined_clean_traits[!(combined_clean_traits$TraitName %in% c("Fruit Mass", "Seed Mass")), ]
```

# Transform plant dataframe from long to wide
```{r}
wide_traits <- spread(combined_clean_traits, key = TraitName, value = TraitValue)

#remove X column from dataframe
wide_plant_traits <- wide_traits[, !colnames(wide_traits) %in% "X", drop = FALSE]
```

```{r}
# Assign data types
wide_plant_traits$`Fruit Length` <- as.numeric(wide_plant_traits$`Fruit Length`)
wide_plant_traits$`Plant Height` <- as.numeric(wide_plant_traits$`Plant Height`)
wide_plant_traits$`Dispersal Syndrome` <- as.factor(wide_plant_traits$`Dispersal Syndrome`)
wide_plant_traits$`Growth Form` <- as.factor(wide_plant_traits$`Growth Form`)
wide_plant_traits$`Fruit Type` <- as.factor(wide_plant_traits$`Fruit Type`)
```

# Summary
```{r}
print("Number of species: ")
length(unique(wide_plant_traits$species))
```

# Write data to csv
```{r}
write.csv(wide_plant_traits, file.path(output_path,"TropicalAndes_all_plant_traits_harmonized.csv"))
write.csv(clean_traits, file.path(output_path,"TropicalAndes_all_plant_traits_harmonized_with_units.csv"))
```

>>>>

---
title: "Subsetting plant species list by fruiting species"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script subsets plant occurrence and trait data by fruiting plant species."
data input: "ADD"
data output: "ADD"
date: "2024-05-29"
output: html_document
---

@@

@@

# Function to get one value per species
```{r}
# categorical data
factor_data_merge <- function(data, traitname) {
# remove duplicate rows
data <- data %>%
  distinct(.keep_all = TRUE)
# remove rows with missing or empty species names
data <- data[!is.na(data$Accepted_species) & data$Accepted_species != "", ]
# Subset data to just species and TraitValue
data <- data[,c("Accepted_species", "TraitValue")]
# Find the most common trait value for each species
  most_common_values <- aggregate(. ~ Accepted_species, data, function(x) {
    # Find the mode (most common value) of the trait values
    mode_value <- names(sort(table(x), decreasing = TRUE))[1]
    return(mode_value)
  })
# Merge most common trait values back to the original data
merged_data <- merge(data, most_common_values, by = "Accepted_species", all.x = TRUE)

@@

# Fruit type
```{r}
traits_fruitType <- traits %>%
  filter(grepl('Fruit type|fruit type|fruit_type', TraitName))
traits_fruitType$OrigTraitValue <- traits_fruitType$TraitValue

@@

```{r}
traits_fruitType %>%
  count(TraitValue)
```

```{r}
# standardize trait values
traits_fruitType <- traits_fruitType %>%
  filter() %>%
  mutate(TraitValue= case_when(
    grepl('Berry|berry|Berry ', OrigTraitValue, ignore.case = TRUE) ~ "berry",
    grepl('capsule|Capsule|Capsule ', OrigTraitValue, ignore.case = TRUE) ~ "capsule",
    grepl('follicle|Follicle', OrigTraitValue, ignore.case = TRUE) ~ "follicle",
  )) %>%
  filter(!is.na(TraitValue))
```

@@
